\chapter{A \langname~fable}
This fable tells of a conflict between the North Wind and the Sun.
\begin{examples}
	\ex
	\script Ramirami: pomarohinang wan eno i Kato Saraw ni wan eno i Toan ni.
	\bits rami^ rami po- maroh -in =ang wan eno i kato saraw ni wan eno i toan ni
	\gloss REDUP before PFV fight AV 3PL.NOM FOC DEM.MED LNK north wind and FOC DEM.MED LNK sun and
	\tr A long time ago, the North Wind and the Sun were fighting.

	\ex
	\script Ihe angoy rahtamanakang kawarewai pa bei doang i soe.
	\bits ihe angoy rah- tama -nak =ang kaw= are wai pa bei doang i soe
	\gloss NEG ABL CAUS to.be.equal PV 3PL.NOM 3PL.POS NMZ think INST more strong LNK who
	\tr They couldn't agree on who was the stronger.

	\ex
	\script Rasiya: yomeninang pa areompor.
	\bits rasiya yomen -in =ang pa are- ompor
	\gloss thus decide AV 3PL.NOM INST NMZ compete
	\tr Thus, they decided on a competition.

	\ex
	\script Daecongket wan sian wan rahsontornak eno i areoysi kesawcow.
	\bits dae- congket wan sian wan rah- sontor -nak eno i are- oysi ke= sawcow
	\gloss IRR win FOC who FOC CAUS remove PV DEM.MED LNK NMZ travel 3SG.POS coat
	\tr The one who caused the traveller to remove his coat would triumph.

	\ex
	\script Pohayo eno Saraw.
	\bits po- hayo eno saraw
	\gloss PFV start DEM.MED wind
	\tr The Wind went first.

	\ex
	\script Kaimkaimong eno i areoysi; ahop komod porabingnak wan kesawcow ong kecoan.
	\bits kaim^ kaim eno i are- oysi ahop komod po- rabing -nak wan ke= sawcow ong ke= coan
	\gloss REDUP blow DEM.MED LNK NMZ travel but to.be.unfortunate PFV to.be.tight PV FOC 3SG.POS coat LOC 3SG.POS body
	\tr She blew and blew at the traveller, but he wrapped his coat around his body even tighter.

	\ex
	\script Rami moay Toan.
	\bits rami moay toan
	\gloss before exist sun
	\tr Then went the Sun.

	\ex
	\script Rahkakaeyong areoysi; tay hayo popospis wan kew.
	\bits rah- ka^ kaey -ong are- oysi tay hayo po^ pospis wan kew
	\gloss CAUS REDUP light LV NMZ travel then start REDUP sweat FOC 3SG.NOM
	\tr She shone brightly on the traveller, and he started to sweat.

	\ex
	\script Ahke: rahokoin keimaw.
	\bits ahke rah- oko -in ke= imaw
	\gloss suddenly CAUS to.be.red AV 3SG.POS face
	\tr And suddenly, his face became red.
\end{examples}